An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-01 uai2017/paper.tex
fix two typos, MC -> MH
All figures should now be up to date.
OK now let me do more changes and then I will tell Xinlei what to do. Almost there ...

% Daniel: use this algorithm style instead
\begin{algorithm}[tb]
\caption{Our acceptance test, {\sc MCminibatch}.}
\label{alg:our_algorithm}
\begin{algorithmic}
    \STATE {\bfseries Input:} number of samples $T$, minibatch size $m$, error
        bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial
        sample $\theta_1$.
    \STATE {\bfseries Output:} a chain of $T$ samples
        $\{\theta_1,\ldots,\theta_T\}$.
    \FOR{$t=1$ {\bfseries to} $T$}
        \STATE -Propose a candidate $\theta'$ from proposal
            $q(\theta' | \theta_t)$.
        \STATE -Draw a minibatch of $m$ points $\{x_1^*,\ldots,x_m^*\}$.
        \STATE -Compute $\Delta^*(\theta_t,\theta')$ and sample variance
            $s^2_{\Delta^*}$.
        \STATE -Estimate moments $\mE[|\Lambda_i-\Lambda|]$ and
            $\mE[|\Lambda_i-\Lambda|^3]$ from the sample, and error $\epsilon$
            from Corollary~\ref{cor:our_bound_delta_prime}.
        \WHILE{$s^2_{\Delta^*} \ge 1$ {\bf or} $\epsilon > \delta$}
            \STATE -Draw $m$ more samples to augment the minibatch, update
                $\Delta^*$, $s^2_{\Delta^*}$ and $\epsilon$ estimates.
        \ENDWHILE
        \STATE -Draw $X_{\rm nc} \sim \mathcal{N}(0,1-s^2_{\Delta^*})$ and
            $X_{\rm corr} \sim C_1(X)$.
    \IF{$\Delta^* + X_{\rm nc} + X_{\rm corr} > 0$}
        \STATE -Accept the candidate, $\theta_{t+1} = \theta'$.
    \ELSE
        \STATE -Reject and re-use the old sample, $\theta_{t+1} = \theta_t$.
    \ENDIF
    \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{corollary}\label{cor:our_bound_delta_prime}
\begin{equation}
\sup_y \left|{\rm Pr}(\Delta^*< y) -\Phi\left(\frac{y-\Delta}{s_{\Delta^*}}\right)\right| \leq \frac{6.4\mE[|X|^3]+2\mE[|X|]}{\sqrt{b}}
\end{equation}
\end{corollary} %\vspace{-10pt}
where the upper bound can be expressed as $\epsilon(\theta,\theta',b)$.
Corollary~\ref{cor:our_bound_delta_prime} shows that the distribution of
$\Delta^*$ approximates a Normal distribution with mean $\Delta$ and variance
$s^2_{\Delta^*}$. Furthermore, it bounds the error with \emph{estimable
quantities}: both $\mE[|X|]$ and $\mE[|X|^3]$ can be estimated as means of
$|\Lambda_i - \Lambda|$ and $|\Lambda_i - \Lambda|^3$, respectively, on each
minibatch. We expect this will often be accurate enough on minibatches with
hundreds of points, but otherwise bootstrap CIs can be computed.

@@

In Sections~\ref{ssec:gaussians} and~\ref{ssec:logistic}, we benchmark our test,
{\sc MHminibatch}, against the test from~\cite{icml2014c1_bardenet14} ({\sc
MHSubLhd}) and from~\cite{cutting_mh_2014}. As highlighted in
Section~\ref{sec:analysis}, the test from~\cite{cutting_mh_2014} can be
implemented using a conservative or a non-conservative variant; we name these
{\sc AustereMH(c)}, and {\sc AustereMH(nc)}, respectively, and they use the
grid-search over minibatch sizes $m$ and per-test thresholds $\epsilon$
described in Appendix~\ref{sssec:grid_search}. Throughout our descriptions, we
refer to a \emph{trial} as the period when an algorithm collects all its desired
samples $\{\theta_1,\ldots,\theta_T\}$, with $T$ usually several thousand.
\textbf{Daniel: better call temperature something other than $T$ ...}

@@

This model is adapted from~\cite{langevin_2011} by increasing the number of
samples to 1 million.  The parameters are $\theta = \langle \theta_1,\theta_2
\rangle$, and the generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i | \theta)$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $T=10,000$.  Taking logs, we get the target as shown
in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.0\linewidth]{logistic_regression_error_regions.png}
	\caption{
    Binary classification accuracy of the MCMC methods on the 1s vs 7s logistic
    regression task for MNIST-13k (left plot) and MNIST-100k (middle plot) as a
    function of cumulative data usage.  The right plot reports performance of
    {\sc MHminibatch} on both datasets when the temperature starts at $T=100$
    and drops to $T=1$ after a ``burn-in'' period of 200 samples of $\theta$.
    For all three plots, one standard deviation is indicated by the shaded error
    regions.
    }
	\label{fig:logistic_performance}
\end{figure*}

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.75\linewidth]{logistic_regression_histograms_mnist.png}
	\caption{
    Minibatch sizes for a representative trial of logistic regression on
    MNIST-13k (analogous to Figure~\ref{fig:gauss_mix_2}). Both axes are on a
    log scale and have the same ranges across the three histograms. See
    Section~\ref{ssec:logistic} for details.
    }
	\label{fig:logistic_histograms_mnist}
\end{figure*}

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.75\linewidth]{logistic_regression_histograms_mnist8m.png}
	\caption{
    Minibatch sizes for a representative trial of logistic regression on
    MNIST-100k (similar to Figure~\ref{fig:logistic_histograms_mnist}). Both
    axes are on a log scale and have the same ranges across the three
    histograms.  See Section~\ref{ssec:logistic} for details.
    }
	\label{fig:logistic_histograms_mnist8m}
\end{figure*}

\begin{table}[t]
\caption{
Average minibatch sizes ($\pm$ one standard deviation) for the methods on the
logistic regression tasks of binary classification on MNIST-13k and MNIST-100k.
The averages are taken over 10 independent trials (5000 samples each) for
MNIST-13k and 5 independent trials (3000 samples each) for MNIST-100k.
}
\small % Daniel: yeah, sorry, but otherwise it doesn't fit in a column
\label{tab:logistic}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Method/Data} & \textbf{MNIST-13k} & \textbf{MNIST-100k}  \\
\hline \\
{\sc MHminibatch}   & $129.8\pm 3.4$    & $202.5 \pm 13.6$ \\
{\sc AustereMH(nc)} & $973.8\pm 49.8$   & N/A \\
{\sc AustereMH(c)}  & $1924.3\pm 52.4$  & $2795.6 \pm 364.0$ \\
{\sc MHSubLhd}      & $10783.4\pm 78.9$ & $14977.3 \pm 582.0$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

For all methods, we impose a uniform prior on $\theta$ and again use a random
walk proposer, with covariance matrix $0.05I$ for MNIST-13k and $0.01I$ for
MNIST-100k. The default temperature setting is a constant at $T=100$ for
MNIST-13k and MNIST-100k. Performance of all methods implicitly relies on the
step size and temperature. Setting temperature too low or step size too high
will generally result in worse performance for all methods. For MNIST-13k, each
method generated 5000 samples for ten independent trials; due to MNIST-100k's
higher computational requirements, the methods generated 3000 samples for five
independent trials.  For additional parameter settings and an investigation on
tuning step sizes, see Appendix~\ref{app:logistic}.

@@

\begin{table}[t]
\caption{Gaussian Mixture Model Statistics}
\small
\label{tab:poissons}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Metric} & {\sc MCminibatch} & {\sc AustereMH} & {\sc MHSubLhd} \\
\hline \\
Equation~\ref{eq:log_prob_poissons} & -1430.0 & -1578.9 & -1232.7 \\
Chi-Squared & 3313.9 & 3647.7 & 2444.1 \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.75\linewidth]{logistic_regression_histograms_mnist8m.png}
	\caption{
    \textbf{Daniel: TODO this is a wrong figure, I'll add what I want later}
    }
	\label{fig:logistic_histograms_appendix}
\end{figure*}

>>>>

% Daniel: use this algorithm style instead
\begin{algorithm}[tb]
\caption{Our acceptance test, {\sc MHminibatch}.}
\label{alg:our_algorithm}
\begin{algorithmic}
    \STATE {\bfseries Input:} number of samples $T$, minibatch size $m$, error
        bound $\delta$, pre-computed correction $C_1(X)$ distribution, initial
        sample $\theta_1$.
    \STATE {\bfseries Output:} a chain of $T$ samples
        $\{\theta_1,\ldots,\theta_T\}$.
    \FOR{$t=1$ {\bfseries to} $T$}
        \STATE -Propose a candidate $\theta'$ from proposal
            $q(\theta' | \theta_t)$.
        \STATE -Draw a minibatch of $m$ points $\{x_1^*,\ldots,x_m^*\}$.
        \STATE -Compute $\Delta^*(\theta_t,\theta')$ and sample variance
            $s^2_{\Delta^*}$.
        \STATE -Estimate moments $\mE[|\Lambda_i-\Lambda|]$ and
            $\mE[|\Lambda_i-\Lambda|^3]$ from the sample, and error $\epsilon$
            from Corollary~\ref{cor:our_bound_delta_prime}.
        \WHILE{$s^2_{\Delta^*} \ge 1$ {\bf or} $\epsilon > \delta$}
            \STATE -Draw $m$ more samples to augment the minibatch, update
                $\Delta^*$, $s^2_{\Delta^*}$ and $\epsilon$ estimates.
        \ENDWHILE
        \STATE -Draw $X_{\rm nc} \sim \mathcal{N}(0,1-s^2_{\Delta^*})$ and
            $X_{\rm corr} \sim C_1(X)$.
    \IF{$\Delta^* + X_{\rm nc} + X_{\rm corr} > 0$}
        \STATE -Accept the candidate, $\theta_{t+1} = \theta'$.
    \ELSE
        \STATE -Reject and re-use the old sample, $\theta_{t+1} = \theta_t$.
    \ENDIF
    \ENDFOR
\end{algorithmic}
\end{algorithm}

@@

\begin{corollary}\label{cor:our_bound_delta_prime}
\begin{equation}
\small
\sup_y \left|{\rm Pr}(\Delta^*< y) -\Phi\left(\frac{y-\Delta}{s_{\Delta^*}}\right)\right| \leq \frac{6.4\mE[|X|^3]+2\mE[|X|]}{\sqrt{b}}
\end{equation}
\end{corollary} %\vspace{-10pt}
where the upper bound can be expressed as $\epsilon(\theta,\theta',b)$.
Corollary~\ref{cor:our_bound_delta_prime} shows that the distribution of
$\Delta^*$ approximates a Normal distribution with mean $\Delta$ and variance
$s^2_{\Delta^*}$. Furthermore, it bounds the error with \emph{estimable
quantities}: both $\mE[|X|]$ and $\mE[|X|^3]$ can be estimated as means of
$|\Lambda_i - \Lambda|$ and $|\Lambda_i - \Lambda|^3$, respectively, on each
minibatch. We expect this will often be accurate enough on minibatches with
hundreds of points, but otherwise bootstrap CIs can be computed.

@@

% Daniel: NOTE, I'm calling temperature "K" so that "T" stays reserved for
% "number of samples of \theta we draw".
In Sections~\ref{ssec:gaussians} and~\ref{ssec:logistic}, we benchmark our test,
{\sc MHminibatch}, against the test from~\cite{icml2014c1_bardenet14} ({\sc
MHSubLhd}) and from~\cite{cutting_mh_2014}. As highlighted in
Section~\ref{sec:analysis}, the test from~\cite{cutting_mh_2014} can be
implemented using a conservative or a non-conservative variant; we name these
{\sc AustereMH(c)}, and {\sc AustereMH(nc)}, respectively, and they use the
grid-search over minibatch sizes $m$ and per-test thresholds $\epsilon$
described in Appendix~\ref{sssec:grid_search}. Throughout our descriptions, we
refer to a \emph{trial} as the period when an algorithm collects all its desired
samples $\{\theta_1,\ldots,\theta_T\}$, generally with $T=3000$ or $T=5000$.

@@

This model is adapted from~\cite{langevin_2011} by increasing the number of
samples to 1 million.  The parameters are $\theta = \langle \theta_1,\theta_2
\rangle$, and the generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i | \theta)$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $K=10,000$.  Taking logs, we get the target as shown
in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.0\linewidth]{logistic_regression_error_regions.png}
	\caption{
    Binary classification accuracy of the MCMC methods on the 1s vs 7s logistic
    regression task for MNIST-13k (left plot) and MNIST-100k (middle plot) as a
    function of cumulative data usage.  The right plot reports performance of
    {\sc MHminibatch} on both datasets when the temperature starts at $K=100$
    and drops to $K=1$ after a ``burn-in'' period of 200 samples of $\theta$.
    For all three plots, one standard deviation is indicated by the shaded error
    regions.
    }
	\label{fig:logistic_performance}
\end{figure*}

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.9\linewidth]{logistic_regression_histograms_mnist.png}
	\caption{
    Minibatch sizes for a representative trial of logistic regression on
    MNIST-13k (analogous to Figure~\ref{fig:gauss_mix_2}). Both axes are on a
    log scale and have the same ranges across the three histograms. See
    Section~\ref{ssec:logistic} for details.
    }
	\label{fig:logistic_histograms_mnist}
\end{figure*}

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.9\linewidth]{logistic_regression_histograms_mnist8m.png}
	\caption{
    Minibatch sizes for a representative trial of logistic regression on
    MNIST-100k (similar to Figure~\ref{fig:logistic_histograms_mnist}). Both
    axes are on a log scale and have the same ranges across the three
    histograms.  See Section~\ref{ssec:logistic} for details.
    }
	\label{fig:logistic_histograms_mnist8m}
\end{figure*}

\begin{table}[t]
\caption{
Average minibatch sizes ($\pm$ one standard deviation) for the methods on the
logistic regression tasks of binary classification on MNIST-13k and MNIST-100k.
The averages are taken over 10 independent trials (5000 samples each) for
MNIST-13k and 5 independent trials (3000 samples each) for MNIST-100k.
}
\small % Daniel: yeah, sorry, but otherwise it doesn't fit in a column
\label{tab:logistic}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Method/Data} & \textbf{MNIST-13k} & \textbf{MNIST-100k}  \\
\hline \\
{\sc MHminibatch}   & $129.8\pm 3.4$    & $202.5 \pm 13.6$ \\
{\sc AustereMH(nc)} & $973.8\pm 49.8$   & $1098.3 \pm 44.9$ \\
{\sc AustereMH(c)}  & $1924.3\pm 52.4$  & $2795.6 \pm 364.0$ \\
{\sc MHSubLhd}      & $10783.4\pm 78.9$ & $14977.3 \pm 582.0$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

For all methods, we impose a uniform prior on $\theta$ and again use a random
walk proposer, with covariance matrix $0.05I$ for MNIST-13k and $0.01I$ for
MNIST-100k. The default temperature setting is a constant at $K=100$ for
MNIST-13k and MNIST-100k. Performance of all methods implicitly relies on the
step size and temperature. Setting temperature too low or step size too high
will generally result in worse performance for all methods. For MNIST-13k, each
method generated 5000 samples for ten independent trials; due to MNIST-100k's
higher computational requirements, the methods generated 3000 samples for five
independent trials.  For additional parameter settings and an investigation on
tuning step sizes, see Appendix~\ref{app:logistic}.

@@

\begin{table}[t]
\caption{Gaussian Mixture Model Statistics}
\small
\label{tab:poissons}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Metric} & {\sc MHminibatch} & {\sc AustereMH} & {\sc MHSubLhd} \\
\hline \\
Equation~\ref{eq:log_prob_poissons} & -1430.0 & -1578.9 & -1232.7 \\
Chi-Squared & 3313.9 & 3647.7 & 2444.1 \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

\begin{figure*}[t]
	\centering
    \includegraphics[width=0.9\linewidth]{logistic_regression_step_size_tests.png}
	\caption{
    \textbf{Daniel: TODO describe}
    }
	\label{fig:logistic_histograms_appendix}
\end{figure*}
